\section*{Acknowledgements}

I would like to thank my supervisor Conor McBride for his contagious
enthusiasm, for his generosity, and most of all for the near limitless
freedom I was given to pursue the problems I was interested in.
%
I learned a lot of the techniques used in this thesis through our
long discussions.
%
Many thanks naturally go to my other co-authors.
%
James Chapman, James McKinna, and Bob Atkey helped me simplify my
generic setup and clarify some of the categorical structures
underlying this work.
%
Andreas Abel, Steven Schäfer, Kathrin Stark, Aliya Hameer,
Alberto Momigliano, and Brigitte Pientka offered,
via the POPLMark Reloaded Challenge, a great opportunity to
compare, contrast, and mutually enrich our respective approaches
to well scoped syntaxes with binding.

I would like to thank my examiners Neel Krishnaswami and Jules Hedges
for the detailed feedback that helped make this thesis a better document,
and for a very pleasant and interesting discussion during the viva.

The results in this thesis could not have been obtained without the
tireless work of the research team behind Agda (Ulf Norell,
Andreas Abel, Nils Anders Danielsson, Jesper Cockx).
%
I have learned so much from using the system, and have had a fantastic
time contributing back to it and collaborating on new features during
the regular Agda Implementors' Meetings.

My work benefited immensely from Matthew Daggit's development and
maintenance work for the Agda standard library.
Our constant discussions on library issues
have deeply informed my own design sensibilities.

This thesis was a long time in the making.
%
I am immensely grateful to both Bart Jacobs and Edwin Brady for trusting
me to conduct exciting postdoctoral research under their supervision
while I had not yet defended.
%
This also means that I have made many wonderful friends in my time in
Strathclyde, Nijmegen, and St Andrews and I want to conclude
by thanking them for making these years very pleasant.

Last but not least, I want to thank my parents who always supported my
choices.
